Removing nonlinear batch effects using adversarial deep neural networks

### 2019-03-05 thesis.tex
Fix citations

To address these issues, we implemented a deep neural network approach to batch correction that works on real-world datasets, and we calculated batch removal measurements based on machine learning algorithms to test for remaining nonlinear effects.
Artificial neural networks are a machine learning tool inspired by the way human brains function; input values pass through layers of linear and nonlinear functions, the final output values are measured against objectives, and the layers of functions are adjusted to bring the outputs closer to the objectives.
This process is repeated until the outputs are sufficiently close to their targets \citep{schmidhuber_deep_2015}.
Autoencoders are a type of neural network that encode and then reconstruct their input, and their traditional objective function is to construct the output as similarly to the input as possible \citep{hinton_reducing_2006}.
Neural networks have historically decreased in effectiveness when working with data from multiple domains \citep{ganin_domain-adversarial_2015}, in part because they may learn based on confounding effects (e.g. which researcher collected the data) instead of learning based on biologically interesting causal effects (e.g. which gene is consistently upregulated in a disease) \citep{louizos_causal_2017}.
Recently, researchers have experimented with discouraging the networks from distinguishing between classes by giving them two objective functions:
1. to learn as much as possible about the input data and
2. to forget any patterns that help distinguish between classes \citep{ganin_domain-adversarial_2015,tzeng_deep_2014}.
This type of dual objective function has been used in conjunction with autoencoders \citep{louizos_variational_2015} where the autoencoder is ``rewarded'' for reconstructing the input faithfully, but ``punished'' for keeping class-specific patterns.
Research has also shown that neural networks are effective in working with biological data; for example, neural networks have been applied to extract biologically relevant latent spaces in RNA-Seq data with as few as 10,000 samples \citep{way_extracting_2017}, and some researchers have used that latent space to generate realistic synthetic biomedical data for other scientific studies \citep{beaulieu-jones_privacy-preserving_2017}.

>>>>

To address these issues, we implemented a deep neural network approach to batch correction that works on real-world datasets, and we calculated batch removal measurements based on machine learning algorithms to test for remaining nonlinear effects.
Artificial neural networks are a machine learning tool inspired by the way human brains function; input values pass through layers of linear and nonlinear functions, the final output values are measured against objectives, and the layers of functions are adjusted to bring the outputs closer to the objectives.
This process is repeated until the outputs are sufficiently close to their targets \citep{schmidhuber_deep_2015}.
Autoencoders are a type of neural network that encode and then reconstruct their input, and their traditional objective function is to construct the output as similarly to the input as possible \citep{hinton_reducing_2006}.
Neural networks have historically decreased in effectiveness when working with data from multiple domains \citep{ganin_domain-adversarial_2015}, in part because they may learn based on confounding effects (e.g. which researcher collected the data) instead of learning based on biologically interesting causal effects (e.g. which gene is consistently upregulated in a disease) \citep{louizos_causal_2017-2}.
Recently, researchers have experimented with discouraging the networks from distinguishing between classes by giving them two objective functions:
1. to learn as much as possible about the input data and
2. to forget any patterns that help distinguish between classes \citep{ganin_domain-adversarial_2015,tzeng_deep_2014-2}.
This type of dual objective function has been used in conjunction with autoencoders \citep{louizos_variational_2015} where the autoencoder is ``rewarded'' for reconstructing the input faithfully, but ``punished'' for keeping class-specific patterns.
Research has also shown that neural networks are effective in working with biological data; for example, neural networks have been applied to extract biologically relevant latent spaces in RNA-Seq data with as few as 10,000 samples \citep{way_extracting_2017}, and some researchers have used that latent space to generate realistic synthetic biomedical data for other scientific studies \citep{beaulieu-jones_privacy-preserving_2017}.
